Ribosome accumulation during early phase resistance training

### 2021-04-12 manuscript.Rmd
added small changes..

---
title: "Ribosome accumulation during early phase resistance training"
author: Daniel Hammarström^1,2,£^&ast;,Sjur J. Øfsteng^2,£^, Nicolai B. Jacobsen^1^,
  Krister B. Flobergseter^1^,  Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
date: ''
output:
  word_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/experimental-physiology.csl
bibliography: resources/bibliography.bib
header-includes: \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

source("./R/libs.R")
# Settings for chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

@@

^1^ Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway.

^2^ Swedish School of Sport and Health Sciences, Stockholm, Sweden.

^3^ Innlandet Hospital Trust, Lillehammer, Norway. </div>

@@

Increased ribosomal density is a phenomena observed in response to resistance training (RT) supporting elevated. A time course of RT-induced accumulation of ribosomes has not been described. We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6) and upstream binding factor (UBF) protein content) in response to twelve RT sessions. Additionally we investigated the effect of fluctuations in training volume on these markers and relationship between RNA accumulation and muscle growth measured as muscle thickness (ultra sound). Eighteen participants were allocated to either a training (TRAIN, *n*=11) or control (CTRL, *n*=7) group. In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3 and 9 sets in sessions 1-4, 5-8 and 9-12, respectively). Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session as well as 48 hrs after session 4, 5, 8, 9 and 12 and after eight days of de-training. Control biopsies were sampled in CTRL at baseline and after 48 hrs and 3-5 weeks. RT led to muscle growth in TRAIN compared to CTRL and concomitant increases in total RNA, rRNA, including precursor rRNA as well as UBF and rpS6 protein through gradual increase throughout the training period, confirmed in comparison to CTRL. Total RNA increased in a curve-linear fashion, most rapidly in response to the first four sessions  (8.6%, 95% CI: [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after Session 8. UBF protein levels explained total RNA levels after controlling for time and increases in total RNA levels predicted RT induced muscle hypertrophy. After de-training, total RNA and specific rRNA species decreased without changes in muscle mass indicating reduced concentrations and biosynthesis of ribosomes. These result underlines a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.

@@

@@

@@

@@

@@

@@

@@

@@

# Figure legends

**Figure 1.** Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).

**Figure 2.** Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF. Black points and error bars represents statistically robust results (a 95% CI not containing 0). C shows western-blots and total protein stains from a representative participant. mRNA data is normalized per total RNA.

**Figure 3.**  Total RNA and ribosomal RNA subspecies in response to trainin. Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A). There were no robust differences between different volume conditions in the training group (A). B shows primer locations targeting different ribosomal RNA subspecies. Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C). Error bars shows 95% CI. Asterisk in C indicates robust differences between volume conditions (a 95% CI of pairwise differences not containing 0).

**Figure 4.** 
Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4). Values are averaged over values from men and women. Individual plots of estimates total RNA increases over time is shown in C together with results from leave-one-out analysis (D). Leave-one-out analysis shows the effect of removing a single participant (black point and error-bars) and individual values from the total RNA per time estimates where green points represents bounds of the 95% CI and yellow points represents mean estimates.

>>>>

---
title: "Ribosome accumulation during early phase resistance training"
author:
  - Daniel Hammarström^1,2,£^*
  - Sjur J. Øfsteng^2,£^
  - Nicolai B. Jacobsen^1^
  - Krister B. Flobergseter^1^
  - Bent R. Rønnestad^1^
  - Stian Ellefsen^1,3^
date: ''
output:
  pdf_document:
    latex_engine: xelatex
  word_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/experimental-physiology.csl
bibliography: resources/bibliography.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

source("./R/libs.R")
library(knitr); library(kableExtra)
# Settings for chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")

@@

^1^ Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway. ^2^ Swedish School of Sport and Health Sciences, Stockholm, Sweden. ^3^ Innlandet Hospital Trust, Lillehammer, Norway. </div>

@@

Increased ribosomal density is a phenomenon observed in response to resistance training (RT) supporting elevated. A time course of RT-induced accumulation of ribosomes has not been described. We mapped markers of ribosomal biogenesis (total RNA, ribosomal RNA (rRNA), ribosomal protein S6 (rpS6), and upstream binding factor (UBF) protein content) in response to twelve RT sessions. Additionally, we investigated the effect of fluctuations in training volume on these markers and the relationship between RNA accumulation and muscle growth measured as muscle thickness (ultrasound). Eighteen participants were allocated to either a training (TRAIN, *n*=11) or a control (CTRL, *n*=7) group. In TRAIN, participants performed unilateral knee extension with either constant (CONST, 6 sets) or a variable volume (VAR, 6, 3, and 9 sets in sessions 1-4, 5-8, and 9-12, respectively). Muscle biopsies were sampled from *m vastus lateralis* in TRAIN before and 48 hrs after the first session and 48 hrs after sessions 4, 5, 8, 9, and 12, and after eight days of de-training. Control biopsies were sampled in CTRL at baseline and after 48 hrs and 3-5 weeks. RT led to muscle growth in TRAIN compared to CTRL and concomitant increases in total RNA, rRNA, including precursor rRNA as well as UBF and rpS6 protein through gradual increase throughout the training period, confirmed in comparison to CTRL. Total RNA increased in a curve-linear fashion, most rapidly in response to the first four sessions  (8.6%, 95% CI: [5.5, 11.7] per session), followed by a plateau and peak values (49.5% [34.2, 66.5] above baseline) after Session 8. UBF protein levels explained total RNA levels after controlling for time, and increases in total RNA levels predicted RT-induced muscle hypertrophy. After de-training, total RNA and specific rRNA species decreased without muscle mass changes indicating reduced concentrations and biosynthesis of ribosomes. These results underline a determinant role for ribosomal biogenesis in RT-induced muscle hypertrophy and that ribosomal biogenesis is sensitive to training cessation.

@@

```{r table1}

# Read dxa data

dxa_whole <- read_excel("./data/tr010_dxa_standarddefinitions.xlsx") %>%
  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
  mutate(value = gsub("kg", "", value), 
         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
         value = as.numeric(value)) %>%

  separate(Etternavn, into = c("Study", "participant")) %>%
  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
  group_by(participant) %>%
  mutate(time = if_else(Date == min(Date), "S0", 
                        if_else(Date == max(Date), "post1w", 
                                "S12")), 
         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
  filter(variable %in% c("HeleBeinmasse", 
                         "HeleFettmasse", 
                         "HeleVevmasse", 
                         "HeleMagermasse" ,
                         "Vekt", 
                         "HeleTotalmasse")) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  mutate(fat_rel = 100*(HeleFettmasse/HeleTotalmasse)) %>%
  dplyr::select(participant, time, Vekt,HeleTotalmasse, fat_rel) %>%
  filter(time == "S0")

# DXA weight stature

read_excel("./data/tr010_age_participant.xlsx") %>%
  inner_join(dxa_whole) %>%
  mutate(age = difftime(recruited, dob)/365.2425, 
         bmi = weight/(height/100)^2) %>%
  dplyr::select(participant, group, height, weight,age, sex,bmi, fat_rel) %>%

  group_by(group, sex) %>%
  summarise(n = n(), 
            Age = mean(age, na.rm = TRUE), 
            age.sd = sd(age, na.rm = TRUE),
            mass = mean(weight), 
            mass.sd = sd(weight),
            stature = mean(height),
            stature.sd = sd(height), 
            BMI = mean(bmi), 
            bmi.sd = sd(bmi), 
            relfat = mean(fat_rel), 
            relfat.sd = mean(fat_rel) ) %>%
  ungroup() %>%
  mutate(n = as.character(n), 
         age = paste0(sprintf("%.1f", round(Age, 1)), " (", sprintf("%.1f", round(age.sd, 1)), ")"),
         mass = paste0(sprintf("%.1f", round(mass, 1)), " (", sprintf("%.1f", round(mass.sd, 1)), ")"),
         stature = paste0(sprintf("%.1f", round(stature, 1)), " (", sprintf("%.1f", round(stature.sd, 1)), ")"),
         BMI = paste0(sprintf("%.1f", round(BMI, 1)), " (", sprintf("%.1f", round(bmi.sd, 1)), ")"),
         fat = paste0(sprintf("%.1f", round(relfat, 1)), " (", sprintf("%.1f", round(relfat.sd, 1)), ")"))%>%
  dplyr::select(group, sex, n, age, mass, stature, BMI, fat) %>%
  pivot_longer(names_to = "variable", values_to = "values", n:fat) %>%
  pivot_wider(names_from = c(sex,group), values_from = values) %>%

  mutate(variable =  factor(variable,
                          levels = c("n", "age", "mass", "stature", "BMI", "fat"), 
                          labels = c("n", "Age (yrs)", "Body mass, (kg)", "Stature (cm)", "Body mass index (kg/m\\textsuperscript{2})", "Body fat (\\%)"))) %>%

  dplyr::select(variable, F_experiment, M_experiment, F_control, M_control) %>%

  kable(caption = "Participant characteristics.", 
        booktabs = TRUE, 
        col.names = c("", "Female", "Male", "Female", "Male"),
        escape = FALSE,
        format = "latex") %>%
  add_header_above(c(" ", "Experimental group" = 2, "Control group" = 2))

@@

```{r primerstable}
# Read qpcr data
qdat <- readRDS("./data/derivedData/qpcr/qpcr_compiled.RDS")

primer_performance <- qdat %>%
  group_by(target) %>%
  summarise(mean.cq = mean(cq, na.rm = TRUE), 
            sd.cq = sd(cq, na.rm = TRUE), 
            mean.eff = mean(eff))

merge_custom <- function(ft, x, columns){
  z <- rle(x)
  rows_at <- cumsum(z$lengths) - z$lengths + 1

  for(i in seq_along(rows_at)){
    for(j in columns)
      ft <- merge_at(x = ft, i = seq( rows_at[i], rows_at[i] + z$lengths[i] - 1), j = j)
  }

  ft
}

# Read primer data 
primer_table <- read_excel("./data/tr010_primers.xlsx", sheet = 1) %>%
  dplyr::select(target = primer_id, forward, reverse, full_name, symbol_ext) %>%
  inner_join(primer_performance) %>%
  mutate(performance = paste0(sprintf("%.1f", round(mean.cq,1)), " (", sprintf("%.1f",        round(sd.cq,1)), "), E = ", sprintf("%.2f", round(mean.eff,2))), 
         forward = paste0("F: 5'-", forward, "-3'"), 
         reverse = paste0("R: 3'-", reverse, "-5'")) %>%

  dplyr::select(symbol_ext, full_name, forward, reverse, performance) %>%
  pivot_longer(names_to = "primer", values_to = "seq", forward:reverse) %>%
dplyr::select(Symbol = symbol_ext, 'Transcript name' = full_name, Sequence = seq, 'Mean Cq (SD) and efficiency' = performance) %>%
    filter(!(Symbol %in% c("RPL32", "MYH7", "MYH2", "MYH1")))

primer_table %>%
  kable(format = "latex", 
        caption = "Primer sequences and average performance.",
        booktabs = TRUE) %>%
  collapse_rows(columns = c(1, 2, 4)) %>%
  landscape()

# 
# flextable() %>%
# merge_custom(x = primer_table$Symbol, columns = c(1, 2, 4)) %>%
# add_header_lines(values = "Table 2. Primer sequences and average performance.") %>%
#     fontsize(size = 8, part = "body") %>%
#   fontsize(size = 10, part = "header") %>%
# autofit()
#

@@

```{r figure1, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Study design (A) observed training loads in response to Constant and Variable volume protocols (B) and training outcomes (muscle strength, C; muscle thickness D).", echo=FALSE}

figure1 <- readRDS("./Figures/rds/figure1.RDS")

figure1

@@

```{r figure2, fig.height=(23*0.75)/2.54, fig.width=8.9/2.54, fig.align="center", fig.cap="Protein (A and B) and mRNA abundances (D and E) of rpS6 and UBF. Black points and error bars represents statistically robust results (a 95\\% CI not containing 0). C shows western-blots and total protein stains from a representative participant. mRNA data is normalized per total RNA.", echo=FALSE}

figure2 <- readRDS("./Figures/rds/figure2.RDS")

figure2

@@

```{r figure3, fig.height=(23*0.75)/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Total RNA and ribosomal RNA subspecies in response to trainin. Total RNA increased compared to non-training controls to post-training (12 sessions) and tended to normalize after de-training (A). There were no robust differences between different volume conditions in the training group (A). B shows primer locations targeting different ribosomal RNA subspecies. Subsets of ribosomal RNA showed robust increases compared to control 48 h after the first session and other after the training period (C). Error bars shows 95\\% CI. Asterisk in C indicates robust differences between volume conditions (a 95\\% CI of pairwise differences not containing 0).", echo=FALSE}

figure3 <- readRDS("./Figures/rds/figure3.RDS")

figure3

```{r table3-regression-models-muscle-thickness}
# Save model

# Model for rpS6 predicting total RNA
comb.s6.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1_s6.RDS")

comb.m1 <- readRDS("./data/derivedData/ubf-tot-rna-model/comb_m1.RDS")

comb.m1 <- summary(comb.m1)
comb.s6.m1 <- summary(comb.s6.m1)

reg_results <- 
  cbind(rbind(comb.m1$fixed,
      comb.m1$random$participant, 
      comb.m1$random$`participant:leg`, 
      comb.m1$spec_pars), 
      data.frame(coef = c("Intercept", 
                          "UBF protein levels (SD-units)", 
                          "Session 1-4\\textsuperscript{b}",
                          "Session 4-8\\textsuperscript{c}", 
                          "Session 8-12\\textsuperscript{d}", 
                          "De-training",
                          "Between Participant variation",
                          "Between Participant:Leg variation", 
                          "Residual SD"))) %>%
  remove_rownames() %>%
  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)

reg_results_s6 <- 
  cbind(rbind(comb.s6.m1$fixed,
      comb.s6.m1$random$participant, 
      comb.s6.m1$random$`participant:leg`, 
      comb.s6.m1$spec_pars), 
      data.frame(coef = c("Intercept", 
                          "rpS6 protein levels (SD-units)", 
                          "Session 1-4\\textsuperscript{b}",
                          "Session 4-8\\textsuperscript{c}", 
                          "Session 8-12\\textsuperscript{d}", 
                          "De-training",
                          "Between Participant variation",
                          "Between Participant:Leg variation", 
                          "Residual SD"))) %>%
  remove_rownames() %>%
  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)

  rbind(reg_results, reg_results_s6) %>%

  mutate(Estimate = round(Estimate, 2), 
         Est.Error = round(Est.Error, 2), 
         lwr = round(lwr, 2), 
         upr = round(upr, 2)) %>%

    kable(col.names = c("Coefficient", 
                        "Estimate\\textsuperscript{a}", 
                        "SD", 
                        "Lower 95\\% CI", 
                        "Upper 95\\% CI"), 
          caption = "Effect of UBF and rpS6 levels, sessions and de-training on RNA-levels.",
          format = "latex", 
          escape = FALSE,
          booktabs = TRUE) %>%
    row_spec(9, hline_after = TRUE) %>%
    footnote(alphabet = c("The dependent variable is total RNA levels (log).",
    "Slope in response to session 1-4.", 
    "Change in slope in session 4-8.",
    "Change in slope in session 8-12."), 
             footnote_as_chunk = TRUE,
             threeparttable = TRUE, 
             escape = FALSE,
             general_title = "")

@@

```{r table4-regression-models-muscle-thickness}
# Save model

hyp.pre.m3 <- readRDS( "./data/derivedData/ubf-tot-rna-model/hyp_pre_m3.RDS")

m3.sum <- summary(hyp.pre.m3)

# combine output from all models

reg_results <- 
  cbind(rbind(m3.sum$fixed,
      m3.sum$random$participant, 
      m3.sum$random$`participant:leg`, 
      m3.sum$spec_pars), 
      data.frame(coef = c("Intercept", 
                          "Sex (Male)", 
                          "Mean Total RNA",
                          "Mean Total RNA increase per session", 
                          "Between Participant variation", 
                          "Between Participant:Leg variation", 
                          "Residual SD"))) %>%
  remove_rownames() %>%
  dplyr::select(coef, Estimate, Est.Error, lwr = `l-95% CI`, upr = `u-95% CI`)

reg_results %>%

  mutate(Estimate = round(Estimate, 2), 
         Est.Error = round(Est.Error, 2), 
         lwr = round(lwr, 2), 
         upr = round(upr, 2)) %>%

      kable(col.names = c("Coefficient", 
                        "Estimate\\textsuperscript{a}", 
                        "SD", 
                        "Lower 95\\% CI", 
                        "Upper 95\\% CI"), 
          caption = "Total RNA as a predictor of muscle growth.",
          format = "latex", 
          escape = FALSE,
          booktabs = TRUE) %>%

   footnote(alphabet = c("The dependent variable is $\\\\Delta$ Muscle thickness (mm)"),

             threeparttable = TRUE, 
             escape = FALSE,
             general_title = "")

@@

```{r figure4, fig.height=17.25/2.54, fig.width=17.8/2.54, fig.align="center", fig.cap="Predictions of muscle thickness increase based on total RNA increases (A) and total RNA abundance (B; see Table 4). Values are averaged over values from men and women. Individual plots of estimates total RNA increases over time is shown in C together with results from leave-one-out analysis (D). Leave-one-out analysis shows the effect of removing a single participant (black point and error-bars) and individual values from the total RNA per time estimates where green points represents bounds of the 95\\% CI and yellow points represents mean estimates.", echo=FALSE}

figure4 <- readRDS("./Figures/rds/figure4_x.RDS")

figure4

@@
